A properly labelled proof system for intuitionistic modal logics

### 2019-09-09 caleidoscope/caleidoscope.tex
started section 3

@@

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuiotionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \qquand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
  \end{equation}
  have already been studied
  in~\cite{Plotkin}.\lutz{check: did they get that wrong in  \cite{Plotkin}?}
\item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer.}\lutz{check terminology} models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as a very unsatisfactory situation\lutz{check what he
    said exactly} (see also \cite{mar:str:indexed}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

@@

This paper is organized as follows, In the next section we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:soundness} and~\ref{sec:completeness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cutelim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture other intuitionistic modal logics. \lutz{what about doing that directly in the system presentation, like Negri does it?}

@@

\section{Intuitionistic modal logics}

	The language of {intuitionisitic modal logic} is the one of intuitionistic propositional logic with the modal operators $\BOX$ and $\DIA$, standing most generally for \emph{necessity} and \emph{possibility}.
	%
	Starting with a set $\mathcal{A}$ of atomic propositions, denoted $a$, modal formulas are constructed from the grammar:
	%
	$$
	A \coloncolonequals
	a \mid A \AND A \mid \TOP \mid A \OR A \mid \BOT \mid A \IMP A \mid \BOX A \mid \DIA A
	$$
	%
	We might sometimes write $\NOT A$ to mean $A \IMP \BOT$.

@@

The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi~\cite{Fischer}.
%, and by Ewald~\cite{Ewald} in the case of intuitionistic tense logic. 
%
It is obtained from intuitionistic propositional logic by adding:
\begin{itemize}
\item the \emph{necessitation rule}: $\BOX A$ is a theorem if $A$ is a theorem; and
\item the following five variants of the \emph{distributivity axiom}:
\begin{equation*}
\label{eq:ik}%\hskip-2em
\begin{array}[t]{r@{\;}l@{\quad}r@{\;}l@{\quad}r@{\;}l}
\kax[1]\colon&\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)
&
\kax[3]\colon&\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)
&
\kax[5]\colon&\DIA\BOT\IMP\BOT
\\
\kax[2]\colon&\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)
&
\kax[4]\colon&(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)\\%x[1ex]
\end{array}
\end{equation*}
\end{itemize}

@@

We write $w \Vdash a$ if $a \in V(w)$, and by definition, we always have $w \Vdash \top$ and never that $w \Vdash \bot$. 
%
Then the relation is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:

$w \Vdash A \AND B$ iff $w \Vdash A$ and $w \Vdash B$

@@

%We write $w \not\Vdash A$  if it is not the case that $w\Vdash A$.\\

We write $w \not\Vdash A$ if it is not the case that $w \Vdash A$, but contrarily to the classical case, we do not have $w \Vdash \neg A$ iff $w \Vdash A$ (since $\neg A$ is defined as $A \SEQ \bot$).

From the monotonicity of the valuation function $V$ , we get a monotonicity property for the relation:
\begin{proposition}(Monotonicity) 
For any formula $A$ and for $w, w' \in W$, if $w \le w'$ and $w \Vdash A$, then $w' \Vdash A$.
\end{proposition}

@@

\todo{Add something else about Fischer-Servi theorem?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}%[h]
	\centering
	\small
	\fbox{
	\begin{minipage}{.95\textwidth}
	\begin{tabular}{@{\!}c@{\quad}c}
		$\vlinf{\rn{id}}{}{\B, x \le y, \Left, \labels{x}{a} \SEQ \Right, \labels{y}{a} }{}$
		&
		$\vlinf{\llabrn\bot}{}{\B, \Left, \labels{x}{\BOT} \SEQ \Right}{}$
		\quad
		$\vlinf{\rlabrn\top}{}{\B, \Left \SEQ \Right, \labels{x}{\TOP}}{}$
		\\\\
		$\vlinf{\llabrn\AND}{}{\B,\Left, \labels{x}{A \AND B} \SEQ \Right}{\B, \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right}$
		&
		$\vliinf{\rlabrn\AND}{}{\B,\Left \SEQ \Right, \labels{x}{A \AND B}}{\B, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left \SEQ \Right, \labels{x}{B}}$
		\\\\
		$\vliinf{\llabrn\OR}{}{\B, \Left, \labels{x}{A \OR B} \SEQ \Right}{\B, \Left, \labels{x}{A} \SEQ \Right}{\B, \Left, \labels{x}{B} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\OR}{}{\B, \Left \SEQ \Right, \labels{x}{A \OR B}}{\B, \Left \SEQ \Right, \labels{x}{A}, \labels{x}{B}}$
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rlabrn\IMP}{\text{\scriptsize $y$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{A \IMP B}}{\B, x \le y, \Left, \labels{y}{A} \SEQ \Right, \labels{y}{B}}$
		}
		\\\\
		\multicolumn{2}{c}{
 		$\vliinf{\llabrn\IMP}{}{\B, x \le y, \Left, \labels{x}{A} \IMP B \SEQ \Right}{\B, x \le y, \labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{y}{A}}{\B, x \le y, \Left, \labels{y}{B} \SEQ \Right}$ 
% 		&$\vliinf{\rlabrn\IMP}{\text{\scriptsize $x \le y \in \B$}}{\B, \Left, \labels{x}{A \IMP B} \SEQ \Right}{\B, \Left \SEQ \Right, \labels{y}{A}}{\B, \Left, \labels{y}{B} \SEQ \Right}$
 		}
		\\\\
		$\vlinf{\llabrn\BOX}{}{\B, x \le y, yRz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, x \le y, yRz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\BOX}{\text{\scriptsize $y, z$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{\BOX A}}{\B, x \le y, y \rel z, \Left \SEQ \Right, \labels{z}{A}}$
		\\\\
		$\vlinf{\llabrn\DIA}{\text{\scriptsize $y$ fresh}}{\B, \Left, \labels{x}{\DIA A} \SEQ \Right}{\B, x \rel y, \Left, \labels{y}{A} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\DIA}{}{\B, x \rel y, \Left \SEQ \Right, \labels{x}{\DIA A}}{\B, x \rel y, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}}$
		\\
		\multicolumn{2}{c}{
		$\mbox{\hbox to .9\linewidth{\dotfill}}$
		}
		\\
		$\vlinf{\rn{refl}}{}{\B, \Left \SEQ \Right}{\B, x\le x, \Left \SEQ \Right}$
		&
		$\vlinf{\rn{trans}}{}{\B, x \le y, y \le z, \Left \SEQ \Right}{\B, x \le y, y \le z, x \le z, \Left \SEQ \Right}$
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rn{F_1}}{\text{\scriptsize $u$ fresh}}{\B, xRy, y \le z, \Left \SEQ \Right}{\B, xRy, y \le z, x \le u, uRz, \Left \SEQ \Right}$
		}
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rn{F_2}}{\text{\scriptsize $u$ fresh}}{\B, xRy,x \le z, \Left \SEQ \Right}{\B, xRy, x \le z, y \le u, zRu, \Left \SEQ \Right }$		
		}
	\end{tabular}		
\end{minipage}
}		
	\caption{System $\labIKp$}
	\label{fig:labIKp}
\end{figure}

@@

\begin{theorem}\label{thm:cutfree-compl}
	%	Let $\CC$ be a set of geometric frame properties as in~\eqref{eq:cla-geometric} and $\labbrn{\CC}$ be the corresponding set of rules following schema~\eqref{eq:modal-grs}.
	%
	For any formula $A$, the following are equivalent.
	%
	\begin{enumerate}
		\item\label{i} $A$ is a theorem of $\IK$ 
		%
		\item\label{ii} $A$ is provable in $\labIKp +\labrn{cut}$ with %\quad
%		
%		\todo{}
		\smash{%\fbox{
			\scalebox{.9}{$\vliinf{\labrn{cut}}{}{\B_1, \B_2, \Left \SEQ \Right}{\B_1, \Left \SEQ \Right, \labels{z}{C}}{\B_2, \Left, \labels{z}{C} \SEQ \Right}$}
		}%}
		%
		\item\label{iii} $A$ is provable in $\labIKp$
		%
		\item\label{iv} $A$ is valid in every birelational frames %satisfying the properties in $\CC$.
	\end{enumerate}
\end{theorem}

@@

\begin{smallequation*}\hspace*{-10em}
	\vlderiibase{\llabrn\IMP}{}{\B_1, \B_2, x \le y, \Left, \labels{x}{A \IMP B} \SEQ \Right}{
		\vliin{\labrn{cut}}{}{\B_1, \B_2, x \le y, \Left, \labels{x}{A \IMP B} \SEQ \Right, \labels{y}{A}}{
			\vlhtr{\DD_1}{\B_1, x \le y, \Left, \labels{x}{A \IMP B} \SEQ \Right, \labels{z}{C}, \labels{y}{A}}
		}{
		\vlhtr{\DD_3^{\rn w}}{\B_2, \Left, \labels{x}{A \IMP B}, \labels{z}{C} \SEQ \Right, \labels{y}{A}}
	}
}{
\vliin{\labrn{cut}}{}{\B_1, \B_2, x \le y, \Left, \labels{y}{B} \SEQ \Right}{
	\vlhtr{\DD_2}{\B_1, x \le y, \Left, \labels{y}{B} \SEQ \Right, \labels{z}{C}}
}{
\vlhtr{\DD_3[v/y]^{\invr{\llabrn\IMP}}}{\B_2, \Left, \labels{y}{B}, \labels{z}{C} \SEQ \Right}
}
}
\end{smallequation*}

@@

\begin{smallequation*}\hspace*{-7em}
	\vlderiibase{\labrn{cut}}{}{\B_1, \B_2, \Left, \labels{x}{\DIA A} \SEQ \Right}{
		\vlin{\llabrn\DIA}{\text{\footnotesize $y'$ fresh}}{\B_1, \Left, \labels{x}{\DIA A} \SEQ \Right, \labels{z}{C}}{
			\vlhtr{\DD_1}{\B_1, x \rel y', \Left, \labels{y'}{A} \SEQ \Right, \labels{z}{C}}
		}
	}{
	\vlhtr{\DD_2}{\B_2, \Left, \labels{x}{\DIA A}, \labels{z}{C} \SEQ \Right}
}
\reducesto
\vlderibase{\llabrn\DIA}{\text{\footnotesize $y''$ fresh (also in $\DD_2$)}}{\B_1, \B_2, \Left, \labels{x}{\DIA A} \SEQ \Right}{
	\vliin{\labrn{cut}}{}{\B_1, \B_2, x \rel y'', \Left, \labels{y''}{A} \SEQ \Right}{
		\vlhtr{\DD_1[y''/y']}{\B_1, x \rel y'', \Left, \labels{y''}{A} \SEQ \Right, \labels{z}{C}}
	}{
	\vlhtr{\DD_2^{\invr{\llabrn\DIA}}}{\B_2, \Left, \labels{y''}{A}, \labels{z}{C} \SEQ \Right}
}
}
\end{smallequation*}

>>>>

\newcommand{\lseq}[3]{#1 , #2 \SEQ #3}

@@

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuiotionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \qquand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
  \end{equation}
  have already been studied
  in~\cite{Plotkin}.\todo{check: did they get that wrong in  \cite{Plotkin}?}
\item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer.}\todo{check terminology in Simpson} models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as a very unsatisfactory situation\todo{check what he
    said exactly} (see also \cite{mar:str:indexed}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

@@

This paper is organized as follows, In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:soundness} and~\ref{sec:completeness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cutelim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture other intuitionistic modal logics. \lutz{what about doing that directly in the system presentation, like Negri does it?}

\todo{maybe paragraph on related work? cite Maffezioli, Naibo and Negri~\cite{Maffezioli}.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Intuitionistic modal logics}\label{sec:intmod}

	The language of {intuitionisitic modal logic} is the one of intuitionistic propositional logic with the modal operators $\BOX$ and $\DIA$, standing most generally for \emph{necessity} and \emph{possibility}.
	%
	Starting with a set $\mathcal{A}$ of atomic propositions, denoted $a$, modal formulas are constructed from the grammar:
	%
	$$
	A \coloncolonequals
	a \mid A \AND A \mid \TOP \mid A \OR A \mid \BOT \mid A \IMP A \mid \BOX A \mid \DIA A
	$$
	%
        \lutz{why do we have $\TOP$?}
	We might sometimes write $\NOT A$ to mean $A \IMP \BOT$. \lutz{do we?}

@@

The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi~\cite{Fischer}.
%, and by Ewald~\cite{Ewald} in the case of intuitionistic tense logic. 
%
It is obtained from intuitionistic propositional logic by adding:
\begin{itemize}
\item the \emph{necessitation rule}: if $A$ is a theorem then $\BOX A$ is also a theorem; and
\item the following five variants of the \emph{distributivity axiom}:
\begin{equation*}
\label{eq:ik}%\hskip-2em
\begin{array}[t]{r@{\;}l@{\quad}r@{\;}l@{\quad}r@{\;}l}
\kax[1]\colon&\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)
&
\kax[3]\colon&\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)
&
\kax[5]\colon&\DIA\BOT\IMP\BOT
\\
\kax[2]\colon&\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)
&
\kax[4]\colon&(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)\\%x[1ex]
\end{array}
\end{equation*}
\end{itemize}

@@

We write $w \Vdash a$ if $a \in V(w)$, and by definition, we always have $w \Vdash \top$ and never that $w \Vdash \bot$. 
%
Then the relation is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:
\begin{equation}
  \label{eq:kripke}
  \parbox{30em}{%
  $w \Vdash A \AND B$ iff $w \Vdash A$ and $w \Vdash B$

@@

%We write $w \not\Vdash A$  if it is not the case that $w\Vdash A$.\\
}
\end{equation}
\todo{fix typography}

We write $w \not\Vdash A$ if it is not the case that $w \Vdash A$, but contrarily to the classical case, we do not have $w \Vdash \neg A$ iff $w \Vdash A$ (since $\neg A$ is defined as $A \IMP \bot$).

From the monotonicity of the valuation function $V$, we get a monotonicity property for the relation:
\begin{proposition}(Monotonicity) 
For any formula $A$ and for $w, w' \in W$, if $w \le w'$ and $w \Vdash A$, then $w' \Vdash A$.
\end{proposition}

@@

\todo{Add something else about Fischer-Servi theorem?}\lutz{no, but we should add the Plotkin-Stirling theorem about the relation between axioms and frame conditions}

\section{The system}\label{sec:system}

In this section we present our properly labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relation atoms, and $\Left$ and $\Right$ are
sets of labelled formulas, all written as list, separated by commas.

Now we can present the inference rules for \emph{system $\labIKp$} for the logic $\IK$.
We obtained this system, shown in Figure~\ref{fig:labIKp}, as follows.
Our starting point was the multiple-conlusion nested sequent system \emph{\`a la} Maehara (as presented in~\cite{kuz:str:maehara}), which can straightforwardly translated into the labelled setting, and which yields the rules $\llabrn\bot$ \todo{$\TOP$ yes/no}, $\llabrn\AND$, $\rlabrn\AND$, $\llabrn\OR$, $\rlabrn\OR$, $\llabrn\DIA$, and $\rlabrn\DIA$ as we show them in Figure~\ref{fig:labIKp}. However, this translation also yields the rules $\rn{id'}$, $\llabrn\IMP'$, and~$\llabrn\BOX'$:
\begin{equation}
  \label{eq:nomon}
  \vlinf{\rn{id}'}{}{\B, \Left, \labels{x}{a} \SEQ \Right, \labels{x}{a} }{}
  \quad
  \vliinf{\llabrn\IMP'}{}{\B,\Left, \labels{x}{A} \IMP B \SEQ \Right}{\B,\labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left, \labels{x}{B} \SEQ \Right}
  \quad
  \vlinf{\llabrn\BOX}{}{\B, xRz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, xRz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}
\end{equation}
that are not sufficient for a complete system, and we will see below why. Before, let us first look at the rules $\rlabrn\IMP$ and $\rlabrn\BOX$. In the  multiple-conlusion nested sequent system of~\cite{kuz:str:maehara}, these are the two rules that force single-conclusion. This is the reason why in our system we need the $\le$ in the relational atoms; we need to go to the future of the current world. 
In the Kripke-semantics in~\eqref{eq:kripke} the two connectives $\IMP$ and $\BOX$ are the ones that need the future-relation $\le$. This relation is reflexive and transitive. In order to capture that in the proof system, we need to add the rules $\rn{refl}$ and $\rn{trans}$. Finally, in the semantics, the two relations $\rel$ and $\le$ are strongly connected through the two conditions $F_1$ and $F_2$ \todo{proper typography}. These need to be reflected in the proof system, which is done by the two rules $\rn{F_1}$ and $\rn{F_2}$.

\todo{monotonicity}

@@

\begin{figure}[!t]
  \begin{center}
	\fbox{
%	\begin{minipage}{.95\textwidth}
	\begin{tabular}{c@{\quad}c}
		$\vlinf{\rn{id}}{}{\B, x \le y, \Left, \labels{x}{a} \SEQ \Right, \labels{y}{a} }{}$
		&
		$\vlinf{\llabrn\bot}{}{\B, \Left, \labels{x}{\BOT} \SEQ \Right}{}$
		\quad
		$\vlinf{\rlabrn\top}{}{\B, \Left \SEQ \Right, \labels{x}{\TOP}}{}$
		\\\\
		$\vlinf{\llabrn\AND}{}{\B,\Left, \labels{x}{A \AND B} \SEQ \Right}{\B, \Left, \labels{x}{A}, \labels{x}{B} \SEQ \Right}$
		&
		$\vliinf{\rlabrn\AND}{}{\B,\Left \SEQ \Right, \labels{x}{A \AND B}}{\B, \Left \SEQ \Right, \labels{x}{A}}{\B, \Left \SEQ \Right, \labels{x}{B}}$
		\\\\
		$\vliinf{\llabrn\OR}{}{\B, \Left, \labels{x}{A \OR B} \SEQ \Right}{\B, \Left, \labels{x}{A} \SEQ \Right}{\B, \Left, \labels{x}{B} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\OR}{}{\B, \Left \SEQ \Right, \labels{x}{A \OR B}}{\B, \Left \SEQ \Right, \labels{x}{A}, \labels{x}{B}}$
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rlabrn\IMP}{\text{\scriptsize $y$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{A \IMP B}}{\B, x \le y, \Left, \labels{y}{A} \SEQ \Right, \labels{y}{B}}$
		}
		\\\\
		\multicolumn{2}{c}{
 		$\vliinf{\llabrn\IMP}{}{\B, x \le y, \Left, \labels{x}{A} \IMP B \SEQ \Right}{\B, x \le y, \labels{x}{A \IMP B}, \Left \SEQ \Right, \labels{y}{A}}{\B, x \le y, \Left, \labels{y}{B} \SEQ \Right}$ 
% 		&$\vliinf{\rlabrn\IMP}{\text{\scriptsize $x \le y \in \B$}}{\B, \Left, \labels{x}{A \IMP B} \SEQ \Right}{\B, \Left \SEQ \Right, \labels{y}{A}}{\B, \Left, \labels{y}{B} \SEQ \Right}$
 		}
		\\\\
		$\vlinf{\llabrn\BOX}{}{\B, x \le y, yRz, \Left, \labels{x}{\BOX A} \SEQ \Right}{\B, x \le y, yRz, \Left, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\BOX}{\text{\scriptsize $y, z$ fresh}}{\B, \Left \SEQ \Right, \labels{x}{\BOX A}}{\B, x \le y, y \rel z, \Left \SEQ \Right, \labels{z}{A}}$
		\\\\
		$\vlinf{\llabrn\DIA}{\text{\scriptsize $y$ fresh}}{\B, \Left, \labels{x}{\DIA A} \SEQ \Right}{\B, x \rel y, \Left, \labels{y}{A} \SEQ \Right}$
		&
		$\vlinf{\rlabrn\DIA}{}{\B, x \rel y, \Left \SEQ \Right, \labels{x}{\DIA A}}{\B, x \rel y, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}}$
		\\
		\multicolumn{2}{c}{
		$\mbox{\hbox to .9\linewidth{\dotfill}}$
		}
		\\
		$\vlinf{\rn{refl}}{}{\B, \Left \SEQ \Right}{\B, x\le x, \Left \SEQ \Right}$
		&
		$\vlinf{\rn{trans}}{}{\B, x \le y, y \le z, \Left \SEQ \Right}{\B, x \le y, y \le z, x \le z, \Left \SEQ \Right}$
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rn{F_1}}{\text{\scriptsize $u$ fresh}}{\B, xRy, y \le z, \Left \SEQ \Right}{\B, xRy, y \le z, x \le u, uRz, \Left \SEQ \Right}$
		}
		\\\\
		\multicolumn{2}{c}{
		$\vlinf{\rn{F_2}}{\text{\scriptsize $u$ fresh}}{\B, xRy,x \le z, \Left \SEQ \Right}{\B, xRy, x \le z, y \le u, zRu, \Left \SEQ \Right }$		
		}
	\end{tabular}		
%\end{minipage}
}		
  \end{center}
  \caption{System $\labIKp$}
	\label{fig:labIKp}
\end{figure}

@@

\begin{theorem}\label{thm:cutfree-compl}
	%	Let $\CC$ be a set of geometric frame properties as in~\eqref{eq:cla-geometric} and $\labbrn{\CC}$ be the corresponding set of rules following schema~\eqref{eq:modal-grs}.
	%
	For any formula $A$, the following are equivalent.
	%
	\begin{enumerate}
		\item\label{i} $A$ is a theorem of $\IK$ 
		%
		\item\label{ii} $A$ is provable in $\labIKp +\labrn{cut}$ with %\quad
%		
%		\todo{}
		\smash{%\fbox{
			\scalebox{.9}{$\vliinf{\labrn{cut}}{}{\B_1, \B_2, \Left \SEQ \Right}{\B_1, \Left \SEQ \Right, \labels{z}{C}}{\B_2, \Left, \labels{z}{C} \SEQ \Right}$}
		}%}
		%

                \lutz{why is that multiplicative on $\B$ and additive on the rest???}
		\item\label{iii} $A$ is provable in $\labIKp$
		%
		\item\label{iv} $A$ is valid in every birelational frames %satisfying the properties in $\CC$.
	\end{enumerate}
\end{theorem}

@@

\begin{smallequation*}\hspace*{-2em}
	\vlderiibase{\llabrn\IMP}{}{\B_1, \B_2, x \le y, \Left, \labels{x}{A \IMP B} \SEQ \Right}{
		\vliin{\labrn{cut}}{}{\B_1, \B_2, x \le y, \Left, \labels{x}{A \IMP B} \SEQ \Right, \labels{y}{A}}{
			\vlhtr{\DD_1}{\B_1, x \le y, \Left, \labels{x}{A \IMP B} \SEQ \Right, \labels{z}{C}, \labels{y}{A}}
		}{
		\vlhtr{\DD_3^{\rn w}}{\B_2, \Left, \labels{x}{A \IMP B}, \labels{z}{C} \SEQ \Right, \labels{y}{A}}
	}
}{
\vliin{\labrn{cut}}{}{\B_1, \B_2, x \le y, \Left, \labels{y}{B} \SEQ \Right}{
	\vlhtr{\DD_2}{\B_1, x \le y, \Left, \labels{y}{B} \SEQ \Right, \labels{z}{C}}
}{
\vlhtr{\DD_3[v/y]^{\invr{\llabrn\IMP}}}{\B_2, \Left, \labels{y}{B}, \labels{z}{C} \SEQ \Right}
}
}
\end{smallequation*}

@@

\begin{smallequation*}\hspace*{-2em}
	\vlderiibase{\labrn{cut}}{}{\B_1, \B_2, \Left, \labels{x}{\DIA A} \SEQ \Right}{
		\vlin{\llabrn\DIA}{\text{\footnotesize $y'$ fresh}}{\B_1, \Left, \labels{x}{\DIA A} \SEQ \Right, \labels{z}{C}}{
			\vlhtr{\DD_1}{\B_1, x \rel y', \Left, \labels{y'}{A} \SEQ \Right, \labels{z}{C}}
		}
	}{
	\vlhtr{\DD_2}{\B_2, \Left, \labels{x}{\DIA A}, \labels{z}{C} \SEQ \Right}
}
\reducesto
\vlderibase{\llabrn\DIA}{\text{\footnotesize $y''$ fresh (also in $\DD_2$)}}{\B_1, \B_2, \Left, \labels{x}{\DIA A} \SEQ \Right}{
	\vliin{\labrn{cut}}{}{\B_1, \B_2, x \rel y'', \Left, \labels{y''}{A} \SEQ \Right}{
		\vlhtr{\DD_1[y''/y']}{\B_1, x \rel y'', \Left, \labels{y''}{A} \SEQ \Right, \labels{z}{C}}
	}{
	\vlhtr{\DD_2^{\invr{\llabrn\DIA}}}{\B_2, \Left, \labels{y''}{A}, \labels{z}{C} \SEQ \Right}
}
}
\end{smallequation*}
